Parameter Space Representation of Pareto Front to Explore Hardware-Software Dependencies

### 2015-01-26 algorithm.tex
AA: rilettura fatta e colpo di aspell

			\nl \\ // Splitting and merging operations\\
			\nl $\psi(\mathscr{R}_{i,h}) \Leftarrow \lbrace 
					\psi(R) | R\in \mathscr{R}_{i,h}	
				\rbrace$; \\
			\nl construct $\mathscr{M}_i$; // the set of pair of contiguous no innovation regions \\
			\nl $\mu(\mathscr{M}_i) = \bigcup_{(R_1,R_2)\in\mathscr{M}_i} \mu(R_1,R_2)$;\\
			\nl\\ // Prepare the configuration space partition for the next iteration \\
			\nl $\mathcal{R}_{i+1} \triangleq
	\psi\left(\mathscr{R}_{i,h}\right) \cup \mathcal{R}_{i,l}\cup
	\left(\mathcal{R}_{i,n}\setminus\bigcup_{\left(R_{1},R_{2}\right)\in\mathcal{C}}\left\{ R_{1},R_{2}\right\} \right)\cup \mu(\mathscr{M}_i)$;\\
		}
	\nl $\tilde{\mathscr{P}} \Leftarrow \mathscr{P}_i$
\end{algorithm}

@@

\subsection{Exploration vs. Exploitation in PS}
The proposed algorithm has been designed in order to be inherently able to balance exploration and exploitation, which are ``two cornerstones of problem solving by search''~\cite{eiben1998evolutionary}. While a detailed survey on the exploration vs. exploitation trade-off can be found in~\cite{vcrepinvsek2013exploration}, here we limitedly recall that \emph{exploration} is the capability of an algorithm to consider different parts of the configuration space, not being confined only to a limited subset and being trapped in local optima. Following the terminology of \cite{weise2012evolutionary}, if the exploration is not sufficiently wide, the algorithm suffers from \emph{premature convergence}, i.e. it starts to analyze only a narrow portion of the configuration space, thus ignoring potential Pareto configurations that are outside this portion. When premature convergence happens, there is insufficient \emph{diversity}~\cite{weise2012evolutionary}, i.e. all the provided solutions are similar.
On the other hand, \emph{exploitation} is the capability to recognize and densely analyze the parts of configuration space that have many Pareto configurations.

While in genetic algorithms exploration and exploration are guaranteed by mutation, crossover and selection, in our algorithm these guarantees are provided for the following reasons:
	\begin{itemize}
		\item Merging and splitting operation are dynamic processes. A portion
of the parameter space may be populated by a large number of small
regions in some eras, and thus a large number of test configurations. However, after having been sufficiently explored, the innovation score of these regions may decrease, they may be merged and therefore that portion of configuration space may be covered by a small number of large regions in next eras. This is a desirable property of PS, since, after a number of eras of intense exploration of a configuration space part, this exploration may turn to be enough thorough and so it may be time for other parts to be minutely scanned. 
PS dynamics can be summarized as follows: while eras alternate, PS zooms out from the configurations subspaces already sufficiently scanned in the past, whereas it zooms in on subspaces which provide novelty (guaranteeing exploitation), meaning that they have not been properly analyzed yet (this helps exploration). 
		\item No portion of parameter space is neglected since, at each era, all the regions, even the least interesting ones, have the chance to be explored with new evaluated points (providing exploration), even though some regions are less densely explored than others.
	\end{itemize}

@@

\subsection{Algorithm Parameters and Scalability} The parameters that
completely determine the algorithm are the configurations budget for each
era $K$ and the threshold factor $\alpha$. From a functional point of
view, $K$ is somewhat similar to the population size of a genetic
approach, determining which configuration should be evaluated in the
subsequent step of the algorithm. However, as seen in the
previous subsection, the way this $K$ configurations are distributed
across the parameters space is completely different.
Authors of \cite{zitzler_ec00} found that
increasing $K$ not necessarily has a considerable impact on the
performance of the algorithm, and this holds also in our algorithm (see for example
Figure~\ref{fig:kimpact} in Subsection~\ref{sec:comparison}).
But while in the genetic approach $K$ must be sufficiently large
in order to guarantee diversity of the Pareto set configurations, in
our algorithm the diversity is not impacted by $K$ since, by
construction, it is guaranteed by the way the configuration space is
partitioned, independently from the number of configurations evaluated at
each iteration.

On the other side, parameter $\alpha$ determines when a region is considered of high
innovation. High values of $\alpha$ mean that a region will hardly be
elected as high-innovation, yielding less splitting operations. This
slows down zooming in on portions of configuration space and makes the
distribution of the simulated configuration more uniform, as there are
fewer small regions with a high test configuration concentration. The
opposite happens when small values of $\alpha$ are used. In this
introductory work we are not explicitly focused on investigating how
these parameters could be fine tuned, nevertheless
we conducted some preliminary tests to confirm the relative stability
of $PS$ with different choices and to select some ``reasonable'' average
values to perform the design space exploration, as we will discuss
in~\secR{ee}.

With regard to the performance of the proposed approach, we observe that the
performance of $PS$ is related to the total number of
different configurations evaluated during the exploration. We can
safely make such assumption, because the process of evaluating a new
configuration is the time-dominating one among all the operations
performed. As we will show in the
next Section, this is especially true in the hardware/software
design space exploration scenario we are addressing, since benchmark
compilation and execution times
are orders of magnitude bigger than the splitting/merging operations and
innovation scores computations.  The $i$-th era would provide a new set
$test_i$ of $K$ configurations to be tested. However, we cannot know
a priori how many of
these will actually be new configurations, i.e. requiring a process of
evaluation. Thus, the number of total new simulations $N_{tot}$ 
is bounded as follows:
\[
K \le N_{tot} \le K \times E_{tot}
\]
where $E_{tot}$ is the number of eras executed and assuming that at
least $K$ evaluation should be performed in order to obtain the first
Pareto $\mathscr{P}_0$.

@@

Another possible approach is closer to the ``Island scheme'' for GA
parallelization~\cite{Borkar14parallel}. The configuration space is
initially divided in $n$ macro-regions, each one assigned to a
different processing node, which autonomously performs all the PS
steps, including merging and splitting, considering the assigned
macro-region as the entire parameter space. Note that, while the
Island scheme for GA parallelization requires a migration operator,
that is not easy to properly tune, we do not have special
requirements. On the other hand, one of the macro-regions may begin to
provide very low innovation at a certain era. In this case, dedicating
to it the same number $K$ of test function as the other macro-regions
is inefficient. To solve this problem, the number of test functions
assigned to each macro-region should be periodically renegotiated, in
favor of the macro-regions providing most innovation.

Opposite to the ``Island  scheme'', a novel parallelization technique
permitted by PS can consist of different instances of PS, each one
running on the whole configuration space but having its own
configuration space partition that separately evolves. Each PS
instance independently performs the zoom-out and zoom-in on
configuration space, splitting and merging its own regions. All PS
instances share the indication of the configurations already tested,
so that a configuration already tested by some PS instance, will not
be tested again by some other PS instance. In addition, PS instances
share the set of non-dominated configuration. Therefore, when a PS
instance evaluates the innovation score of its regions (steps
\ref{pers02.enu:novelty_score_of_a_configuration} and
\ref{enu:region_innovation_score} of \secR{steps}), it considers the
solutions found by the other PS instances. In this way, the
configuration space is spanned in parallel by multiple PS instances,
but, although autonomous, they are not isolated; on the contrary, they
continuously exchange information and influence each other, thanks to
sharing the already simulated configurations and the non-dominated
one. Otherwise stated, all the PS instances (independently but
jointly) contribute at the formation of the Pareto-set.

>>>>

			\nl \\ // Splitting and merging operations\\
			\nl $\psi(\mathscr{R}_{i,h}) \Leftarrow \lbrace 
					\psi(R) | R\in \mathscr{R}_{i,h}	
				\rbrace$; \\
			\nl construct $\mathscr{M}_i$; // the set of pair of contiguous no innovation regions \\
			\nl $\mu(\mathscr{M}_i) = \bigcup_{(R_1,R_2)\in\mathscr{M}_i} \mu(R_1,R_2)$;\\
			\nl\\ // Prepare the configuration space partition for the next iteration \\
			\nl $\mathcal{R}_{i+1} \triangleq
	\psi\left(\mathscr{R}_{i,h}\right) \cup \mathcal{R}_{i,l}\cup
	\left(\mathcal{R}_{i,n}\setminus\bigcup_{\left(R_{1},R_{2}\right)\in\mathscr{M}_i }\left\{ R_{1},R_{2}\right\} \right)\cup \mu(\mathscr{M}_i)$;\\
		}
	\nl $\tilde{\mathscr{P}} \Leftarrow \mathscr{P}_i$
\end{algorithm}

@@

\subsection{Exploration vs. Exploitation in PS}
The proposed algorithm has been designed in order to be inherently able to balance exploration and exploitation, which are ``two cornerstones of problem solving by search''~\cite{eiben1998evolutionary}. While a detailed survey on the exploration vs. exploitation trade-off can be found in~\cite{vcrepinvsek2013exploration}, here we limitedly recall that \emph{exploration} is the capability of an algorithm to consider different parts of the configuration space, not being confined only to a limited subset, and thus not being trapped in local optima. Following the terminology of \cite{weise2012evolutionary}, if the exploration is not sufficiently wide, the algorithm suffers from \emph{premature convergence}, i.e. it starts to analyze only a narrow portion of the configuration space, thus ignoring potential Pareto configurations that are outside this portion. When premature convergence happens, there is insufficient \emph{diversity}~\cite{weise2012evolutionary}, i.e. all the provided solutions are similar.
On the other hand, \emph{exploitation} is the capability to recognize and densely analyze the parts of configuration space that have many Pareto configurations.

While in genetic algorithms exploration and exploitation are guaranteed by mutation, crossover and selection, in our algorithm these guarantees are provided for the following reasons:
	\begin{itemize}
		\item Merging and splitting operations are dynamic processes. A portion
of the parameter space may be populated by a large number of small
regions in some eras, and thus a large number of test configurations. However, after having been sufficiently explored, the innovation score of these regions may decrease, they may be merged and therefore that portion of configuration space may be covered by a small number of large regions in next eras. This is a desirable property of PS, since, after a number of eras of intense exploration of a configuration space part, this exploration may turn out to be enough thorough and so it may be time for other parts to be minutely scanned. 
PS dynamics can be summarized as follows: while eras alternate, PS zooms out from the configurations subspaces already sufficiently scanned in the past, whereas it zooms in on subspaces which provide novelty (guaranteeing exploitation), meaning that they have not been properly analyzed yet (this helps exploration). 
		\item No portion of parameter space is neglected since, at each era, all the regions, even the least interesting ones, have the chance to be explored with new evaluated points (providing exploration), even though some regions are less densely explored than others.
	\end{itemize}

@@

\subsection{Algorithm Parameters and Scalability} The parameters that determine the algorithm are the configurations budget for each
era $K$ and the threshold factor $\alpha$. From a functional point of
view, $K$ is somewhat similar to the population size of a genetic
approach, determining how many configurations should be evaluated in the
subsequent step of the algorithm. However, as seen in the
previous subsection, the way this $K$ configurations are distributed
across the parameters space is completely different.
Authors of \cite{zitzler_ec00} found that
increasing $K$ not necessarily has a considerable impact on the
performance of the algorithm, and this holds also in our algorithm (see for example
Figure~\ref{fig:kimpact} and Subsection~\ref{sec:comparison}).
Nonetheless, while in the genetic approach $K$ must be sufficiently large
in order to guarantee the diversity of the Pareto set configurations, in
our algorithm the diversity is not impacted by $K$ since, by
construction, it is guaranteed by the way the configuration space is
partitioned, independently of the number of configurations evaluated at
each iteration.

On the other side, parameter $\alpha$ determines when a region is considered of high
innovation. High values of $\alpha$ mean that a region will hardly be
elected as high-innovation, producing less splitting operations. This
slows down zooming in on portions of configuration space and makes the
distribution of the simulated configurations more uniform, since there are
fewer small regions with a high test configuration concentration. The
opposite happens when small values of $\alpha$ are used. In this
introductory work we are not explicitly focused on investigating how
these parameters could be fine tuned, nevertheless
we conducted some preliminary tests to confirm the relative stability
of $PS$ with different choices and to select some ``reasonable'' average
values to perform the design space exploration, as we will discuss
in~\secR{ee}.

With regard to the performance of the proposed approach, we observe it is related to the total number of
different configurations evaluated during the exploration. We can
safely make such assumption, because the process of evaluating a new
configuration is the time-dominating one among all the operations
performed. As we will show in the
next Section, this is especially true in the hardware/software
design space exploration scenario we are addressing, since benchmark
compilation and execution times
are orders of magnitude bigger than the splitting/merging operations and
innovation scores computations.  The $i$-th era would provide a new set
$test_i$ of $K$ configurations to be tested. However, we cannot know
a priori how many of
these will actually be new configurations, i.e. requiring a process of
evaluation. Thus, the number of total new simulations $N_{tot}$ 
is bounded as follows:
\[
K \le N_{tot} \le K \times E_{tot}
\]
where $E_{tot}$ is the number of eras executed and assuming that at
least $K$ evaluation should be performed in order to obtain the first
Pareto $\mathscr{P}_0$.

@@

Another possible approach is closer to the ``Island scheme'' for GA
parallelization~\cite{Borkar14parallel}. The configuration space is
initially divided in $n$ macro-regions, each one assigned to a
different processing node, which autonomously performs all the PS
steps. When merging and splitting, each processing node consider the assigned
macro-region as the entire parameter space. Note that, while the
Island scheme for GA parallelization requires a migration operator,
that is not easy to properly tune, we do not have special
requirements. On the other hand, one of the macro-regions may begin to
provide very low innovation at a certain era. In this case, dedicating
to it the same number $K$ of test functions as to the other macro-regions
is inefficient. To solve this problem, the number of test functions
assigned to each macro-region should be periodically renegotiated, in
favor of the macro-regions providing most innovation.

Opposite to the ``Island  scheme'', a novel parallelization technique
permitted by PS can consist of different instances of PS, each one
running on the whole configuration space but having its own
configuration space partition that separately evolves. Each PS
instance independently performs the zoom-out and zoom-in on
configuration space, splitting and merging its own regions. All PS
instances share the indication of the configurations already tested,
so that a configuration already tested by some PS instance, will not
be tested again by some other PS instance. In addition, PS instances
share the set of non-dominated configurations. Therefore, when a PS
instance evaluates the innovation score of its regions (steps
\ref{pers02.enu:novelty_score_of_a_configuration} and
\ref{enu:region_innovation_score} of \secR{steps}), it considers also the
solutions found by the other PS instances. In this way, the
configuration space is spanned in parallel by multiple PS instances,
that, although autonomous, are not isolated; on the contrary, they
continuously exchange information and influence each other, thanks to
the sharing of the the set of already simulated configurations and the set of non-dominated
ones. Otherwise stated, all the PS instances (independently but
jointly) contribute to the formation of the Pareto-set.

### 2015-01-26 experiments.tex
AA: rilettura fatta e colpo di aspell

\section{A Case Study: VLIW Hardware/Software Design Space}
\secL{ee}
Although PS has been formally introduced and described as a general
multi-objective optimization algorithm, in this work we will
explicitly refer to the more restricted (but still very large)
embedded system design scenario,  since its properties
are the one that suggested to the authors the concepts of Pareto-based
innovation and region scoring system.

In particular, we will use a parametrized Very Long Instruction
Word (VLIW) architecture~\cite{kathail_tr00} as testbed. The philosophy
behind a VLIW system makes this a natural choice for several reasons:
\begin{itemize}
\item \emph{Multi-Objective trade-offs}: VLIW architectures allow designers
to trade between power, energy and performances objectives, resulting
in extended Pareto sets which are an ideal ground for our testing
purposes focused on parameter-space representation
\item \emph{Software Compiler awareness}: code scheduling for the execution of
applications is statically obtained by the compiler which
performs code transformations ``being aware'' of the underlying hardware
configuration.  This tight hardware/software dependence makes the
VLIW scenario perfectly suitable for investigating hardly predictable
parameter interactions.
\end{itemize}

@@

\subsection{Design Space: Hardware and Software Parameters}
\secL{designspace}
Conceptually, system parameters can be classified in three main categories:
\emph{processor}, \emph{memory sub-system} and \emph{compiler}. The
first two categories are directly related to the physical
implementation of the system, so parameters of those
classes are usually referred as ``hardware parameters''. On the other
side, all the parameters affecting the behavior of the compiler
are referred as ``software parameters'', being directly involved in
the process of generating the application executable running on the
underlying hardware.

Processor hardware parameters are directly related to the VLIW core
and includes the size of the register files and the number of
functional units of each type. On the software side, a set of the most
impacting code compilation parameters has been selected.
Table~\ref{tab:param} show the complete set of parameters included in
the design space and their possible values. The total size of design
space, which grows with the product of parameter cardinalities, is
about $1.33 \times 10^{14}$.

As regards the class of benchmark being considered, a set applications representative of some common
frequently running code kernels has been selected. Table~\ref{tab:bench} shows the set of applications
along with a brief description.
% IEEE table style
%\begin{table}
%	\centering
%	\caption{Benchmarks}
%	\label{tab:bench}
%	\begin{tabular}{ll}
%	\hline
%	\multicolumn{1}{c}{Benchmark} & \multicolumn{1}{c}{Application} \\
%	\hline
%	Alloca\_test & Memory array allocation test \\
%	Bmm & Matrices multiplication and elements sum \\
%	Fir\_int & Finite impulse response \\
%	Mm & Floating point matrices multiplication \\
%	Sha256 & Cryptocurrency header Hashing \\
%	Wave & Wavefront computation \\
%	\hline
%	\end{tabular}
%\end{table}

@@

To evaluate and compare the performance indexes of different
architectures for a specific application, one needs to simulate the
architecture running the code of the application. To make
architectural exploration possible both the compiler and the simulator
have to be retargetable. The open source project Trimaran~\cite{trimaran_hp} provides these
tools and together with estimation framework \ee~\cite{palpatti_estimedia03} has been
adopted as a tested and flexible platform for carrying out the
experiments described in the previous subsection.

We will now show a functional scheme highlighting the main blocks of
the \ee{} framework and the interface with the Trimaran tools. The
input for the whole evaluation flow is the source of the application
involved in the optimization and the configuration of the architecture
being evaluated.  With reference to Figure~\ref{fig:evaluation_flow}
this input is represented by the blocks \emph{App.c} and
\emph{Config}.
\begin{figure}
        \centering
        \includegraphics[width=0.65\textwidth]{pictures/evaluation_flow.eps}
        \caption{Block diagram of the framework.}
        \label{fig:evaluation_flow}
\end{figure}

@@

The results presented in the next section compare PS with one widely
adopted Multi-objective Genetic Algorithm~\cite{knowles_techrep06}. To
the best of our knowledge, genetic approaches in general still
represent the best compromise between efficiency (i.e. time required
for exploration) and accuracy of the reported Pareto Sets.  Note that,
for the reasons described above, mono-objective approaches have been
explicitly excluded from this analysis.  We also exclude other
approaches (e.g.  dependency based~\cite{givargis_tvlsi02}) that make
use of some a-priori knowledge about the role and meaning of each
parameter, since our aim is evaluate the capacity of focusing on
interesting parameter regions without any external judgment or help
based on heuristics.  In particular in this work when referring to
genetic approach (or GA) we will denote the widely spread 
\emph{SPEA}~\cite{zitzler_eurogen01} variant of the Multi-objective
Genetic Algorithm, remanding to
works like~\cite{zitzler_ec00}~\cite{zitzler_tec03} for a detailed
analysis on how it could differ from other genetic approaches and what
could be expected from similar multi-objective strategies.

@@

The set of default algorithm tuning parameters used for both the
genetic and PS approaches to perform the simulations are shown in
Table~\ref{tab:defaults}. Given the similarity, both in terms of
meaning and functionality, of the parameter $K$ with the population size
of the genetic approach, they both have been set to the same values. As discussed
when introducing the main PS algorithm concepts, a sensitivity analysis on the algorithm parameters is beyond the scope of this work. Nevertheless, we conducted several tests with
different values of $K$, similar to the ones shown in
Figure~\ref{fig:kimpact} which refers to the alloca\_test application
and a total budget of 500 configurations. We found that initially increasing values
of $K$ seems to improve both Pareto (as in the picture on the left) and objective
distribution (on the right), but after some threshold value the $K$ does
not seem to provide any benefit, even returning to slightly worse
results for the last value of $K$ considered (100). Similar trends have been
found for the other benchmarks (not included for space reasons),
showing on average a value of $K$ around 30 as
reasonable sweet spot for the objectives. A complete resource with the
whole set of results and data obtained in the tests performed can be found online
at~\cite{ps_results}.

@@

\subsection{Evaluation metrics}
\secL{metrics}
A qualitative evaluation of Paretos is probably the simplest way for
having a quick glance on algorithm behavior and results. However, in
tests involving a comparison between different exploration strategies
this approach could not help in capturing the whole picture, missing
some important properties related to the distribution of the solutions
found.
In order to evaluate the Paretos from a quantitative point-of-view we
will consider two metrics, namely, the \emph{variation range} and the
\emph{average normalized absolute dispersion error}. Both of them permit to quantify the extent of the Pareto front that is an important factor to consider when evaluating the quality of a Pareto front. Indeed, authors in \cite{zitzler_ec00,weise2012evolutionary} observe that an exploration algorithm should preferably provide Pareto fronts whose objective function range is large. In other words, following the terminology of~\cite{weise2012evolutionary}, a \emph{uniform convergence} should be guaranteed.
Pareto fronts with large objective function ranges permit to investigate in a broader way the potentials of the designed device. This is important, in general, since if only a narrow portion of the Pareto front is provided due to non-uniform convergence, the decision maker may neglect some important options~\cite{weise2012evolutionary}. Having a wide Pareto front becomes essential, in particular, when designing a device that may be required to satisfy different constraints. For example, the decision maker may want to design two different multi-purpose processors. The first design may adapt to situations in which the processor is used for real time elaboration (thus requiring low execution time) and is attached to the powerline (thus not having stringent power dissipation constraints). The second design, on the contrary, may fit situations in which the energy is provided by a battery (stringent power constraints) but only background computations are performed (long execution time is tolerable). These two processors may be successfully designed maintaining the same architecture, using different values for the design parameters. A wide Pareto front helps in achieving this goal.

@@

In this Section we first analyze the results of GA and PS approaches from a
qualitative perspective, showing how Pareto curves compare when
plotted on the bidimensional space of power and performance
objectives. Next, we perform a quantitative
analysis of the results in order to more accurately figure out in which
features/properties the two approaches can differentiate.

\figR{pareto_fronts_100} and \figR{pareto_fronts_500} show the
resulting Pareto fronts for two scenarios of 100 and 500
simulations budget respectively. The set of algorithm parameters involved and their
values are the one previously shown in Table~\ref{tab:defaults}. It
should be pointed out again that the number of generations (or eras)
has been chosen so that only the given amount of different simulations is
actually performed. Thus, when accounting these simulations only
unique simulations have been considered, removing repetitions and
unfeasible configurations.

@@

Low budget Pareto sets are shown in \figR{pareto_fronts_100}. Note
that with a so small number of configuration being investigated the
two sets are often overlapping, being also strongly dependent upon the
application being considered.  These results are obtained in about 5
eras and demonstrate that with a so limited budget PS can compete with
genetic approach while even discovering very different Pareto fronts.
Considering the extended simulation budget as depicted in
\figR{pareto_fronts_500}, we can qualitatively observe that even if PS
does not strictly outperforms genetic in terms of dominance, it shows
in some cases largely Pareto in terms of extension. We can intuitively
relate this to the ``novelty-based'' score system, so that instead of
focusing on getting better points inside a given range it tries to
enlarge the range itself, resulting in less dense but more extended
Paretos. Of course we cannot classify this different behavior as
being strictly better or worse when compared to the genetic one: other
design factors should be evaluated on a case-by-case basis, e.g. the
desired granularity of the results, objective range constraints or
estimated error of the models used.

To better evaluate results from a different perspective, we will
adopted the two quantitative metrics of variation range and average absolute
dispersion error, which have been described and motivated
in~\secR{metrics}. In particular, a comparison between the variation
ranges of both power dissipation and execution time, obtained by PS
and GA for different benchmarks is shown in \figR{dispersion}(a) and
(b) respectively.  As it can be observed, the PS exploration provides
solutions which fall on a range that is, on average, wider than the
one provided by a GA exploration for power dissipation and execution
time, respectively.
%\begin{table}
%  \centering
%  \begin{tabular}{c}
%    \includegraphics[width=0.7\textwidth]{pictures/variation_power.eps} \\
%    (a) \\
%    \includegraphics[width=0.7\textwidth]{pictures/variation_etime.eps} \\
%    (b) \\
%    \includegraphics[width=0.7\textwidth]{pictures/dispersion.eps} \\
%    (c) 
%  \end{tabular}
%  \caption{(a)(b) Variation range obtained by PS and GA for different benchmarks
%  and average dispersion absolute dispersion (c)}
%  \label{fig:dispersion}
%\end{table}

@@

\figR{dispersion}(c) shows the average normalized absolute dispersion
errors for different benchmarks for PS and GA exploration.  As it can
be observed, GA shows a tendency in discovering more uniformly
dispersed points (i.e. higher values in the chart). These results,
together with the variation ranges of \figR{dispersion}(a) and (b)
confirm the of density vs. extension trade-offs when comparing PS and
GA approaches. We highlight that PS shows a stronger \emph{uniform convergence} (discussed in \secR{metrics}), i.e. it provides a more extended Pareto front, avoiding the designer to neglect some important options, and offering a wider view of the potentiality of the designed device.

>>>>

\section{A Case Study: VLIW Hardware/Software Design Space}
\secL{ee}
Although PS has been formally introduced and described as a general
multi-objective optimization algorithm, in this work we will
explicitly refer to the more restricted (but still very large)
embedded system design scenario,  since its properties
are the ones that suggested to the authors the concepts of Pareto-based
innovation and region scoring system.

In particular, we use a parametrized Very Long Instruction
Word (VLIW) architecture~\cite{kathail_tr00} as testbed. The philosophy
behind a VLIW system makes this a natural choice for several reasons:
\begin{itemize}
\item \emph{Multi-Objective trade-offs}: VLIW architectures allow designers
to balance power, energy and performances objectives, resulting
in extended Pareto sets which are an ideal ground for our testing
purposes focused on parameter-space representation
\item \emph{Software Compiler awareness}: code scheduling for the execution of
applications is statically obtained by the compiler which
performs code transformations ``being aware'' of the underlying hardware
configuration.  This tight hardware/software dependence makes the
VLIW scenario perfectly suitable for investigating hardly predictable
parameter interactions.
\end{itemize}

@@

\subsection{Design Space: Hardware and Software Parameters}
\secL{designspace}
Conceptually, system parameters can be classified in three main categories:
\emph{processor}, \emph{memory sub-system} and \emph{compiler}. The
first two categories are directly related to the physical
implementation of the system, and thus parameters of those
classes are usually referred as ``hardware parameters''. On the other
side, all the parameters affecting the behavior of the compiler
are referred as ``software parameters'', being directly involved in
the process of generating the application executable running on the
underlying hardware.

Processor hardware parameters are directly related to the VLIW core
and include the size of the register files and the number of
functional units of each type. On the software side, a set of the most
impacting code compilation parameters has been selected.
Table~\ref{tab:param} shows the complete set of parameters included in
the design space and their possible values. The total size of the design
space, which grows with the product of parameter cardinalities, is
about $1.33 \times 10^{14}$.

As regards the class of benchmark being considered, a set of applications representative of some common
frequently running code kernels has been selected. Table~\ref{tab:bench} shows the set of applications
along with a brief description.
% IEEE table style
%\begin{table}
%	\centering
%	\caption{Benchmarks}
%	\label{tab:bench}
%	\begin{tabular}{ll}
%	\hline
%	\multicolumn{1}{c}{Benchmark} & \multicolumn{1}{c}{Application} \\
%	\hline
%	Alloca\_test & Memory array allocation test \\
%	Bmm & Matrices multiplication and elements sum \\
%	Fir\_int & Finite impulse response \\
%	Mm & Floating point matrices multiplication \\
%	Sha256 & Cryptocurrency header Hashing \\
%	Wave & Wavefront computation \\
%	\hline
%	\end{tabular}
%\end{table}

@@

To evaluate and compare the performance indexes of different
architectures for a specific application, one needs to simulate the
architecture running the code of the application. To make
architectural exploration possible both the compiler and the simulator
have to be retargetable. The open source project Trimaran~\cite{trimaran_hp} provides these
tools and, together with the estimation framework \ee~\cite{palpatti_estimedia03}, has been
adopted as a tested and flexible platform for carrying out the
experiments.

We will now show a functional scheme highlighting the main blocks of
the \ee{} framework and the interface with the Trimaran tools. The
input for the whole evaluation flow is the source file of an application
benchmark and the configuration of the architecture
being evaluated.  With reference to Figure~\ref{fig:evaluation_flow}
this input is represented by the blocks \emph{App.c} and
\emph{Config}.
\begin{figure}
        \centering
        \includegraphics[width=0.65\textwidth]{pictures/evaluation_flow.eps}
        \caption{Block diagram of the framework.}
        \label{fig:evaluation_flow}
\end{figure}

@@

The results presented in the next section compare PS with one widely
adopted Multi-objective Genetic Algorithm~\cite{knowles_techrep06}. To
the best of our knowledge, genetic approaches in general still
represent the best compromise between efficiency (i.e. time required
for exploration) and accuracy of the reported Pareto Sets.  Note that,
for the reasons described above, mono-objective approaches have been
explicitly excluded from this analysis.  We also exclude other
approaches (e.g.  dependency based~\cite{givargis_tvlsi02}) that make
use of some a-priori knowledge about the role and the semantics of each
parameter, since our aim is evaluate the capacity of focusing on
interesting parameter regions without any external judgment or help
based on heuristics.  In this work, when referring to
genetic approach (or GA), we consider the widely spread 
\emph{SPEA2}~\cite{zitzler_eurogen01} variant of the Multi-objective
Genetic Algorithm, remanding to
works like~\cite{zitzler_ec00}~\cite{zitzler_tec03} for a detailed
analysis on how it could differ from other genetic approaches and what
could be expected from similar multi-objective strategies.

@@

The set of default algorithm tuning parameters used for both the
genetic and PS approaches is shown in
Table~\ref{tab:defaults}. Given the similarity, both in terms of
meaning and functionality, between the parameter $K$ and the population size
of the genetic approach, they both have been set to the same values. As discussed
when introducing the main PS algorithm concepts, a sensitivity analysis on the algorithm parameters is beyond the scope of this work. Nevertheless, we conducted several tests with
different values of $K$, similar to the ones shown in
Figure~\ref{fig:kimpact}, which refers to the alloca\_test application
and a total budget of 500 configurations. We found that initially increasing values
of $K$ seems to improve both Pareto front (as in the picture on the left) and objective
distribution (on the right), but, after some threshold value, $K$ does
not seem to provide any benefit, even returning to slightly worse
results for the last values of $K$ considered (100). Similar trends have been
found for the other benchmarks (not included for space reasons),
showing on average a value of $K$ around 30 as
reasonable sweet spot for the objectives. A complete resource with the
whole set of results and data obtained in the tests performed can be found online
at~\cite{ps_results}.

@@

\subsection{Evaluation metrics}
\secL{metrics}
A qualitative evaluation of Paretos is probably the simplest way to
have a quick glance on algorithm behavior and results. However, in
tests involving a comparison between different exploration strategies,
this approach could not help in capturing the whole picture, missing
some important properties related to the distribution of the solutions
found.
In order to evaluate the Paretos from a quantitative point-of-view we
will consider two metrics, namely, the \emph{variation range} and the
\emph{average normalized absolute dispersion error}. Both of them permit to quantify the extent of the Pareto front that is an important factor to consider when evaluating its quality. Indeed, authors of \cite{zitzler_ec00,weise2012evolutionary} observe that an exploration algorithm should preferably provide Pareto fronts whose objective function range is large. In other words, following the terminology of~\cite{weise2012evolutionary}, a \emph{uniform convergence} should be guaranteed.
Pareto fronts with large objective function ranges permit to investigate in a broader way the potentials of the designed device. This is important, in general, since if only a narrow portion of the Pareto front is provided due to non-uniform convergence, the decision maker may neglect some important options~\cite{weise2012evolutionary}. Having a wide Pareto front becomes essential, in particular, when designing a device that may be required to satisfy different constraints. For example, the decision maker may want to design two different multi-purpose processors. The first design may adapt to situations in which the processor is used for real time elaboration (thus requiring low execution time) and is attached to the powerline (thus not having stringent power dissipation constraints). The second design, on the contrary, may fit situations in which the energy is provided by a battery (stringent power constraints) but only background computations are performed (long execution time is tolerable). These two processors may be successfully designed maintaining the same architecture and simply using different values for the design parameters. A wide Pareto front helps in achieving this goal.

@@

In this Section we first analyze the results of GA and PS approaches from a
qualitative perspective, showing how Pareto curves compare when
plotted on the bidimensional space of power and performance
objectives. Next, we perform a quantitative
analysis of the results in order to more accurately figure out in which
features/properties the two approaches differentiate.

\figR{pareto_fronts_100} and \figR{pareto_fronts_500} show the
resulting Pareto fronts for two scenarios of 100 and 500
simulation budget, respectively. The set of algorithm parameters involved and their
values are the ones previously shown in Table~\ref{tab:defaults}. It
should be pointed out again that the number of generations (or eras)
has been chosen so that only the given amount of different simulations is
actually performed. Thus, when accounting these simulations only
unique simulations have been considered, removing repetitions and
unfeasible configurations.

@@

Low budget Pareto sets are shown in \figR{pareto_fronts_100}. Note
that with such a small number of configurations being investigated, the
two sets are often overlapping, being also strongly dependent upon the
application being considered.  These results are obtained in about 5
eras and demonstrate that, with such a limited budget, PS can compete with
genetic approach while even discovering very different Pareto fronts.
Considering the extended simulation budget as depicted in
\figR{pareto_fronts_500}, we can qualitatively observe that even if PS
does not strictly outperforms GA in terms of dominance, it shows
in some cases larger Pareto in terms of extension. We can intuitively
relate this to the ``novelty-based'' score system, so that instead of
focusing on getting better points inside a given range PS tries to
enlarge the range itself, resulting in less dense but more extended
Paretos. Of course we cannot classify this different behavior as
being strictly better or worse when compared to the genetic one: other
design factors should be evaluated on a case-by-case basis, e.g. the
desired granularity of the results, objective range constraints or
estimated error of the adopted models.

To better evaluate results from a different perspective, we
adopt the \emph{variation range} and the \emph{average absolute
dispersion error}, the two quantitative metrics described and motivated
in~\secR{metrics}. In particular, a comparison between the variation
ranges of both power dissipation and execution time, obtained by PS
and GA for different benchmarks, is shown in \figR{dispersion}(a) and~(b), respectively.  As it can be observed, the PS exploration provides
solutions which fall on a range that is, on average, wider than the
one provided by a GA exploration for power dissipation and execution
time, respectively.
%\begin{table}
%  \centering
%  \begin{tabular}{c}
%    \includegraphics[width=0.7\textwidth]{pictures/variation_power.eps} \\
%    (a) \\
%    \includegraphics[width=0.7\textwidth]{pictures/variation_etime.eps} \\
%    (b) \\
%    \includegraphics[width=0.7\textwidth]{pictures/dispersion.eps} \\
%    (c) 
%  \end{tabular}
%  \caption{(a)(b) Variation range obtained by PS and GA for different benchmarks
%  and average dispersion absolute dispersion (c)}
%  \label{fig:dispersion}
%\end{table}

@@

\figR{dispersion}(c) shows the average normalized absolute dispersion
errors for different benchmarks for PS and GA exploration.  As it can
be observed, GA shows a tendency in discovering more uniformly
dispersed points (i.e. higher values in the chart). These results,
together with the variation ranges of \figR{dispersion}(a) and~(b)
confirm the density vs. extension trade-offs when comparing PS and
GA approaches. We highlight that PS shows a stronger \emph{uniform convergence} (discussed in \secR{metrics}), i.e. it provides a more extended Pareto front, avoiding the designer to neglect some important options, and offering a wider view of the potentials of the designed device.
